G\"odel's God in Isabelle/HOL

### 2013-11-08 Formalizations/Isabelle/GoedelGodSession/output/document/GoedelGod.tex
several updates wrt to tpi script and Isabelle session
Improve LaTeX

%
\begin{isabellebody}%
\def\isabellecontext{GoedelGod}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isamarkupsection{Introduction%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
A formalization and (partial) automation of Dana Scott's version \cite{ScottNotes}
 of Goedel's ontological argument \cite{GoedelNotes} in quantified modal logic KB (QML KB) is
 presented. QML KB is in turn modeled as a fragment of classical higher-order logic (HOL). 
 Thus, the formalization is essentially a formalization in HOL. The employed embedding 
 of QML KB in HOL is adapting the work of Benzm\"uller and Paulson \cite{J23,B9}.
 Note that the QML KB formalization employs quantification over individuals and 
 quantification over sets of individuals (poperties).

 The formalization presented here has been carried out and verified in the Isabelle/HOL 
 proof assistant; for more information on this system see the textbook by Nipkow, 
 Paulson, and Wenzel \cite{Isabelle}. More recent tutorials on Isabelle can be found 
 at: \url{http://isabelle.in.tum.de}.

 Some further notes: \sloppy
 \begin{enumerate}
 \item This LaTeX text document has been produced automatically from the Isabelle source
 code document at 
 \url{https://github.com/FormalTheology/GoedelGod/tree/master/Formalizations/Isabelle/GoedelGodSession} 
 with the Isabelle build tool.
 \item The formalization presented here is related to the THF \cite{J22} and 
    Coq \cite{Coq} formalizations at
    \url{https://github.com/FormalTheology/GoedelGod/tree/master/Formalizations/}.
 \item All reasoning gaps in Scott's proof script have been automated 
    with Sledgehammer \cite{Sledgehammer}, performing remote calls to the higher-order automated
    theorem prover LEO-II \cite{LEO-II}. These calls suggest the 
    Metis \cite{Metis} calls as given below. The Metis proofs are verified in Isabelle/HOL.
 \item For consistency checking, the model finder Nitpick \cite{Nitpick} has been employed.

@@

The types \isa{i} for possible worlds (or states) and $\mu$ for individuals 

@@

QML formulas are identified with certain HOL terms of type \isa{i\ {\isasymRightarrow}\ bool}. 
This type will be abbreviated in the remainder as \isa{{\isasymsigma}}.%

@@

\isa{{\isasymforall}}, \isa{{\isasymPi}}, and \isa{{\isasymexists}}. Further connectives could be introduced analogously. 

@@

\ mforall{\isacharunderscore}indset\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}{\isacharparenleft}{\isasymmu}\ {\isasymRightarrow}\ {\isasymsigma}{\isacharparenright}\ {\isasymRightarrow}\ {\isasymsigma}{\isacharparenright}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ {\isacharparenleft}{\isachardoublequoteopen}{\isasymPi}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where}\ {\isachardoublequoteopen}{\isasymPi}\ P\ {\isasymequiv}\ {\isacharparenleft}{\isasymlambda}w{\isachardot}\ {\isasymforall}x{\isachardot}\ P\ x\ w{\isacharparenright}{\isachardoublequoteclose}\isanewline

@@

\ mbox\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymsigma}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ {\isacharparenleft}{\isachardoublequoteopen}{\isasymbox}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where}\ {\isachardoublequoteopen}{\isasymbox}\ {\isasymphi}\ {\isasymequiv}\ {\isacharparenleft}{\isasymlambda}w{\isachardot}\ {\isasymforall}v{\isachardot}\ {\isasymnot}\ w\ r\ v\ {\isasymor}\ {\isasymphi}\ v{\isacharparenright}{\isachardoublequoteclose}\isanewline

@@

The meaning of \isa{P} is restricted by axioms \isa{A{\isadigit{1}}{\isacharparenleft}a{\isacharslash}b{\isacharparenright}}: $\all \phi 
[P(\neg \phi) \biimp \neg P(\phi)]$ (Either a property or its negation is positive, but not both.) 
and \isa{A{\isadigit{2}}}: $\all \phi \all \psi [(P(\phi) \wedge \nec \all x [\phi(x) \imp \psi(x)]) 

@@

\ \ \ \ A{\isadigit{1}}a{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymPhi}{\isachardot}\ P\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ m{\isasymnot}\ {\isacharparenleft}{\isasymPhi}\ x{\isacharparenright}{\isacharparenright}\ m{\isasymRightarrow}\ m{\isasymnot}\ {\isacharparenleft}P\ {\isasymPhi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ A{\isadigit{1}}b{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymPhi}{\isachardot}\ m{\isasymnot}\ {\isacharparenleft}P\ {\isasymPhi}{\isacharparenright}\ m{\isasymRightarrow}\ P\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ m{\isasymnot}\ {\isacharparenleft}{\isasymPhi}\ x{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ A{\isadigit{2}}{\isacharcolon}\ \ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymPhi}{\isachardot}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasympsi}{\isachardot}\ {\isacharparenleft}P\ {\isasymPhi}\ m{\isasymand}\ {\isasymbox}\ {\isacharparenleft}{\isasymforall}\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ {\isasymPhi}\ x\ m{\isasymRightarrow}\ {\isasympsi}\ x{\isacharparenright}{\isacharparenright}{\isacharparenright}\ m{\isasymRightarrow}\ P\ {\isasympsi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}%

@@

properties are possibly exemplified). T1 is proved directly by Sledghammer with command \isa{sledgehammer\ {\isacharbrackleft}provers\ {\isacharequal}\ remote{\isacharunderscore}leo{\isadigit{2}}{\isacharbrackright}}. This successful attempt then suggests to 
instead try the Metis call in the line below. This Metis call generates a proof object 

@@

\ T{\isadigit{1}}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymPhi}{\isachardot}\ P\ {\isasymPhi}\ m{\isasymRightarrow}\ {\isasymdiamond}\ {\isacharparenleft}{\isasymexists}\ {\isasymPhi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}\ \ \isanewline

@@

as $G(x) \biimp \forall \phi [P(\phi) \to \phi(x)]$ (A God-like being possesses 

@@

\ G\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymmu}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ \isakeyword{where}\ {\isachardoublequoteopen}G\ {\isacharequal}\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymPhi}{\isachardot}\ P\ {\isasymPhi}\ m{\isasymRightarrow}\ {\isasymPhi}\ x{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}%

@@

\ \isakeyword{where}\ A{\isadigit{4}}{\isacharcolon}\ \ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymPhi}{\isachardot}\ P\ {\isasymPhi}\ m{\isasymRightarrow}\ {\isasymbox}\ {\isacharparenleft}P\ {\isasymPhi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}%

@@

$\ess{\phi}{x} \biimp \phi(x) \wedge \all \psi (\psi(x) \imp \nec \all y (\phi(y) 

@@

\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ ess\ x\ {\isacharequal}\ {\isasymPhi}\ x\ m{\isasymand}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasympsi}{\isachardot}\ {\isasympsi}\ x\ m{\isasymRightarrow}\ {\isasymbox}\ {\isacharparenleft}{\isasymforall}\ {\isacharparenleft}{\isasymlambda}y{\isachardot}\ {\isasymPhi}\ y\ m{\isasymRightarrow}\ {\isasympsi}\ y{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}%

@@

defined as $\NE(x) \biimp \all \phi [\ess{\phi}{x} \imp \nec \ex y \phi(y)]$ (Necessary 

@@

\ NE\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymmu}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ \isakeyword{where}\ {\isachardoublequoteopen}NE\ {\isacharequal}\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymPhi}{\isachardot}\ {\isasymPhi}\ ess\ x\ m{\isasymRightarrow}\ {\isasymbox}\ {\isacharparenleft}{\isasymexists}\ {\isasymPhi}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}%

@@

@@

\paragraph{Acknowledgments:} Nik Sultana, Jasmin Blanchette and Larry Paulson provided 
very important help wrt consistency checking in Isabelle. Jasmin Blanchette instructed us on how to 
produce latex documents from Isabelle sources, and he showed us useful tricks in Isabelle.%

>>>>

%
\begin{isabellebody}%
\def\isabellecontext{GoedelGod}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isamarkupsection{Introduction%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
Dana Scott's version \cite{ScottNotes}
 of Goedel's ontological argument \cite{GoedelNotes} for God's existence is here
 formalized in quantified modal logic KB (QML KB) within the proof assistant Isabelle/HOL. 
 QML KB is  modeled as a fragment of classical higher-order logic (HOL); 
 thus, the formalization is essentially a formalization in HOL. The employed embedding 
 of QML KB in HOL is adapting the work of Benzm\"uller and Paulson \cite{J23,B9}.
 Note that the QML KB formalization employs quantification over individuals and 
 quantification over sets of individuals (properties).

 The gaps in Scott's proof have been automated 
 with Sledgehammer \cite{Sledgehammer}, performing remote calls to the higher-order automated
 theorem prover LEO-II \cite{LEO-II}. Sledgehammer then suggests the 
 Metis \cite{Metis} calls. The Metis proofs are verified by Isabelle/HOL.
 For consistency checking, the model finder Nitpick \cite{Nitpick} has been employed.

 Isabelle is described in the textbook by Nipkow, 
 Paulson, and Wenzel \cite{Isabelle} and in tutorials available 
 at: \url{http://isabelle.in.tum.de}.

\subsection{Related Work}

 The formalization presented here is related to the THF \cite{J22} and 
 Coq \cite{Coq} formalizations at 
 \url{https://github.com/FormalTheology/GoedelGod/tree/master/Formalizations/}.

 A medieval ontological argument by Anselm was formalized in PVS by John Rushby \cite{ToDo}.

@@

The types \isa{i} for possible worlds and $\mu$ for individuals 

@@

QML formulas are translated as HOL terms of type \isa{i\ {\isasymRightarrow}\ bool}. 
This type is abbreviated as \isa{{\isasymsigma}}.%

@@

\isa{{\isasymforall}}, \isa{{\isasymPi}}, and \isa{{\isasymexists}}. Other connectives could be introduced analogously. 

@@

\ mforall{\isacharunderscore}indset\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}{\isacharparenleft}{\isasymmu}\ {\isasymRightarrow}\ {\isasymsigma}{\isacharparenright}\ {\isasymRightarrow}\ {\isasymsigma}{\isacharparenright}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ {\isacharparenleft}{\isachardoublequoteopen}{\isasymPi}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where}\ {\isachardoublequoteopen}{\isasymPi}\ {\isasymPhi}\ {\isasymequiv}\ {\isacharparenleft}{\isasymlambda}w{\isachardot}\ {\isasymforall}x{\isachardot}\ {\isasymPhi}\ x\ w{\isacharparenright}{\isachardoublequoteclose}\isanewline

@@

\ mbox\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymsigma}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ {\isacharparenleft}{\isachardoublequoteopen}{\isasymbox}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where}\ {\isachardoublequoteopen}{\isasymbox}\ {\isasymphi}\ {\isasymequiv}\ {\isacharparenleft}{\isasymlambda}w{\isachardot}\ {\isasymforall}v{\isachardot}\ \ w\ r\ v\ {\isasymlongrightarrow}\ {\isasymphi}\ v{\isacharparenright}{\isachardoublequoteclose}\isanewline

@@

The meaning of \isa{P} is restricted by axioms \isa{A{\isadigit{1}}{\isacharparenleft}a{\isacharslash}b{\isacharparenright}}: $\all \varphi 
[P(\neg \varphi) \biimp \neg P(\varphi)]$ (Either a property or its negation is positive, but not both.) 
and \isa{A{\isadigit{2}}}: $\all \varphi \all \psi [(P(\varphi) \wedge \nec \all x [\varphi(x) \imp \psi(x)]) 

@@

\ \ \ \ A{\isadigit{1}}a{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymphi}{\isachardot}\ P\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ m{\isasymnot}\ {\isacharparenleft}{\isasymphi}\ x{\isacharparenright}{\isacharparenright}\ m{\isasymRightarrow}\ m{\isasymnot}\ {\isacharparenleft}P\ {\isasymphi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ A{\isadigit{1}}b{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymphi}{\isachardot}\ m{\isasymnot}\ {\isacharparenleft}P\ {\isasymphi}{\isacharparenright}\ m{\isasymRightarrow}\ P\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ m{\isasymnot}\ {\isacharparenleft}{\isasymphi}\ x{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ A{\isadigit{2}}{\isacharcolon}\ \ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymphi}{\isachardot}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasympsi}{\isachardot}\ {\isacharparenleft}P\ {\isasymphi}\ m{\isasymand}\ {\isasymbox}\ {\isacharparenleft}{\isasymforall}\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ {\isasymphi}\ x\ m{\isasymRightarrow}\ {\isasympsi}\ x{\isacharparenright}{\isacharparenright}{\isacharparenright}\ m{\isasymRightarrow}\ P\ {\isasympsi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}%

@@

properties are possibly exemplified). T1 is proved directly by Sledgehammer with command \isa{sledgehammer\ {\isacharbrackleft}provers\ {\isacharequal}\ remote{\isacharunderscore}leo{\isadigit{2}}{\isacharbrackright}}. 
Sledgehammer suggests to call Metis with axioms A1a and A2. 
Metis sucesfully generates a proof object 

@@

\ T{\isadigit{1}}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymphi}{\isachardot}\ P\ {\isasymphi}\ m{\isasymRightarrow}\ {\isasymdiamond}\ {\isacharparenleft}{\isasymexists}\ {\isasymphi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}\ \ \isanewline

@@

as $G(x) \biimp \forall \varphi [P(\phi) \to \varphi(x)]$ \\ (A God-like being possesses 

@@

\ G\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymmu}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ \isakeyword{where}\ {\isachardoublequoteopen}G\ {\isacharequal}\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymphi}{\isachardot}\ P\ {\isasymphi}\ m{\isasymRightarrow}\ {\isasymphi}\ x{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}%

@@

\ \isakeyword{where}\ A{\isadigit{4}}{\isacharcolon}\ \ {\isachardoublequoteopen}{\isacharbrackleft}{\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymphi}{\isachardot}\ P\ {\isasymphi}\ m{\isasymRightarrow}\ {\isasymbox}\ {\isacharparenleft}P\ {\isasymphi}{\isacharparenright}{\isacharparenright}{\isacharbrackright}{\isachardoublequoteclose}%

@@

$\ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi (\psi(x) \imp \nec \all y (\varphi(y) 

@@

\ \ \ \ {\isachardoublequoteopen}{\isasymphi}\ ess\ x\ {\isacharequal}\ {\isasymphi}\ x\ m{\isasymand}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasympsi}{\isachardot}\ {\isasympsi}\ x\ m{\isasymRightarrow}\ {\isasymbox}\ {\isacharparenleft}{\isasymforall}\ {\isacharparenleft}{\isasymlambda}y{\isachardot}\ {\isasymphi}\ y\ m{\isasymRightarrow}\ {\isasympsi}\ y{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}%

@@

defined as $\NE(x) \biimp \all \varphi [\ess{\varphi}{x} \imp \nec \ex y \varphi(y)]$ (Necessary 

@@

\ NE\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymmu}\ {\isasymRightarrow}\ {\isasymsigma}{\isachardoublequoteclose}\ \isakeyword{where}\ {\isachardoublequoteopen}NE\ {\isacharequal}\ {\isacharparenleft}{\isasymlambda}x{\isachardot}\ {\isasymPi}\ {\isacharparenleft}{\isasymlambda}{\isasymphi}{\isachardot}\ {\isasymphi}\ ess\ x\ m{\isasymRightarrow}\ {\isasymbox}\ {\isacharparenleft}{\isasymexists}\ {\isasymphi}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}%

@@

\ {\isacharbrackleft}provers\ {\isacharequal}\ remote{\isacharunderscore}leo{\isadigit{2}}{\isacharbrackright}{\isacharparenleft}T{\isadigit{1}}\ T{\isadigit{3}}\ G{\isacharunderscore}def\ sym{\isacharparenright}%

@@
